---
title: "Detailed Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detailed Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  dpi = 300,
  dev = "png",        # Switch to png
  fig.path = "figures/", # Save figures in a local folder per vignette
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)


library(knitr)
```


<!-- badges: start -->

<!-- badges: end -->

`SummArIzeR` is an R
package, that allows an
easy use of `EnrichR` to
compare enrichment
results from multiple
databases of multiple
conditions. It results in
a clustering of enriched
terms and enables the
annotation of these
terms by creating a
promt for large language
models such as gpt4.
Results can be
visualized in a Bubbleplot or Heatmap.



## Example

This is an example which shows you how to run `SummArIzeR`. We
create an example dataframe from two celltypes and two conditions:

```{r, warning = FALSE,message = FALSE, example}
library(SummArIzeR)
library(enrichR)
library(factoextra)

set.seed(1234)

genelist_df <- data.frame(
  genes = c(
    "IL6", "IFNG", "IL10", "TNF", "CXCL8", "STAT3", "STAT1", "JAK1", "JAK2", "SOCS1", 
    "SOCS3", "IL1B", "IL18", "IL12A", "IL23A", "GATA3", "RORC", "CCR7", "CXCR3", "CCL2",
    "CCL5", "IRF4", "IRF5", "IL4R", "IL2", "FOXP3", "TGFB1", "IL7R", "IL15", "NFKB1", 
    "IL21", "CXCR5", "TYK2", "IL22", "IL17A", "IL17F", "MCP1", "IL1RN", "CXCL10", "CXCL11", 
    "CD40LG", "IL6R", "IL27", "CD28", "CD80", "CD86", "IL2RA", "CTLA4", "PDCD1"
  ),
  Comparison = c(
    "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", 
    "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", 
    "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", 
    "Disease1_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease1_vs_Control", 
    "Disease1_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease1_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control", "Disease2_vs_Control"
  ),
  CellType = c(
    "CD4", "CD4", "CD4", "Monocytes", "Monocytes", "CD4", "CD4", "Monocytes", "CD4", "Monocytes", 
    "CD4", "Monocytes", "Monocytes", "CD4", "CD4", "CD4", "CD4", "CD4", "CD4", "Monocytes", 
    "Monocytes", "CD4", "Monocytes", "CD4", "CD4", "CD4", "Monocytes", "CD4", "Monocytes", "Monocytes", 
    "CD4", "CD4", "Monocytes", "CD4", "CD4", "CD4", "Monocytes", "Monocytes", "CD4", "CD4", 
    "CD4", "Monocytes", "CD4", "CD4", "Monocytes", "Monocytes", "CD4", "CD4", "CD4"
  ),
  log2fold = c(1.65922417, 1.74830165, -0.85544186, 1.32179050, 0.56698208, 0.07638380, 0.94635326, -1.46133361, 0.62796916, 0.82025914, -0.16903290, 0.87644901, 1.73868899, -0.97828470, -0.15082871, 1.76005809, 1.91290571, -1.53005055, -0.10001167, 0.24133098, 1.61612555, -1.44515933, 1.95556692, 1.78667293, -1.67024977, 0.05684714, -0.43918613, 1.62295252, -0.21212149, 1.34401704, 0.95038247, 1.24422057, -0.44756687, 0.74067892, -1.98420664, 1.33166432, -1.97066341, -1.16936411, 1.62640563, 0.44711457, -0.48176304, -0.25691366, -1.85027587, 1.89415966, -0.27299500, 1.83030639, 1.55101962, 0.55991508, 1.88386644)
)
```

We can perform enrichment analysis across the different conditions: CellType and Comparison.
In this example, we extract the **top 5 enriched terms** from each database and for every condition.

The input is a data.frame containing gene-level data, with the following required columns:

`genes`: character vector of gene symbols.
`conditions`: condition labels (e.g., comparisons or cell types).
`log2fold` (optional): used when splitting by regulation. 

If `split_by_reg = TRUE` (default is `FALSE`), enrichment is performed separately for up- and down-regulated genes.

The `logFC_threshold` parameter (default: `0`) defines the log2 fold-change cutoff to distinguish up/down regulation when using `split_by_reg`.

To filter enriched terms, an adjusted p-value `pval_threshold` cutoff is applied.  

Additional key parameters include:
`n` (default: 10): number of top enriched terms returned per condition and per category.
`min_genes_threshold` (default: 3): minimum number of overlapping genes required for a term to be considered in downstream clustering and annotation.


The enrichment `categories` can be chosen from the set of libraries available in
[EnrichR](https://maayanlab.cloud/Enrichr/#libraries):

```{r, enrichment, message = FALSE, cache=TRUE, results='hide'}
Termlist_all<-extractMultipleTerms(genelist_df, condition_col = c("CellType", "Comparison"), 
                                  categories = c("GO_Biological_Process_2023",
                                                 "Reactome_2022", "BioPlanet_2019"), 
                                  pval_threshold = 0.05, n = 5, split_by_reg = T)

```
Enrichment analysis can also be performed offline by setting the 'enrichment_method' to '"gmt"'.
One or more local '.gmt' libraries must be provided via a file path.

Libraries can be downloaded from : 
[EnrichR](https://maayanlab.cloud/Enrichr/#libraries)
If the format is '.txt', the file can be renamed to '.gmt'.

```{r, enrichment_offline, message = FALSE, cache=TRUE, results='hide'}
library_path<-c(
  "../examples/libraries/GO_Biological_Process_2023.gmt",
  "../examples/libraries/Reactome_2022.gmt",
"../examples/libraries/BioPlanet_2019.gmt"
)
Termlist_all_offline<-extractMultipleTerms(genelist_df, 
                                           condition_col = c("CellType", "Comparison"),
                                          gmt_files = library_path , pval_threshold = 0.05, n = 5,
                                          split_by_reg = T,enrichment_method = "gmt" )

```
```{r, view, include=T}
head(Termlist_all, n = 5)
head(Termlist_all_offline, n = 5)
```

Edges below the
similarity threshold `ts`
are deleted. The optimal
threshold can be
validated by checking
the number of clusters,
the connected terms and
the modularity. The
threshold value
determines which edges
between terms (nodes)
are retained in the
network (those with a
distance weight higher
than `ts`). Smaller values
include more edges;
larger values result in
sparser graphs:

```{r, evaluate_cluster, include=T, fig.width=1, fig.height=1}
evaluateThreshold(Termlist_all)
```
The similarity between the different terms is visualized in a heatmap, while their clustering relationships can be explored through a dendrogram.



```{r, cluster_network, include=T, fig.width=15, fig.height=12}
plot<-TRUplotIgraph(Termlist_all, ts  = 0.3)
#htmltools::tagList(plot)
```
The network can also be
interactively
visualized, and if
needed, the threshold
can be readjusted.

```{r, plot_network, fig.width=1, fig.height=1}
plot
```
Adjusting the threshold `ts`  controls the number of terms grouped into each cluster.

```{r, cluster_network_dif_ts, eval=FALSE}
plot<-TRUplotIgraph(Termlist_all, ts  = 0.5)
#htmltools::tagList(plot)
```
```{r, cluster_network_dif_ts_code, include=FALSE}
plot<-TRUplotIgraph(Termlist_all, ts  = 0.5)
#htmltools::tagList(plot)
```

If we increase the threshold `ts`, we increase the number of cluster:
```{r, plot_cluster_network_dif_ts, include=T, fig.width=1, fig.height=1}
plot
#htmltools::tagList(plot)
```
After treshold
adjustment, clusters can
be assigned to the
dataframe and the Prompt
can be generated:

```{r, prompt, include=T}
Genelist_test_cluster<-returnIgraphCluster(Termlist_all, ts = 0.3)
head(Genelist_test_cluster, n = 5)
generateGPTPrompt(Genelist_test_cluster)
```

The prompt can be
queried using a LLM like
ChatGPT. The result can
be copied into R. Now, a
final data frame
containing the cluster
annotations can be
created:

```{r cluster_annotation, message=FALSE, warning= FALSE}
#Enter vector from LLM 
cluster_summary <- c(
  '1' = 'Cytokine and Interleukin Signaling in Immune Regulation',
  '2' = 'T Cell Activation and Differentiation',
  '3' = 'Regulation of T Cell Transcription Factors',
  '4' = 'Regulation of Inflammatory Cytokine Responses',
  '5' = 'T Helper Cell Differentiation and Inflammatory Responses',
  '6' = 'Innate Immune and Inflammatory Response to Pathogens',
  '7' = 'Regulation of Protein Localization to the Nucleus',
  '8' = 'Positive Regulation of Interleukin Production',
  '9' = 'Chemokine Receptor Expression During T Cell Polarization'
)

print(cluster_summary)

```

The clusters can now be
annotated based on the
vectors defined above.


```{r, annotation, warning = FALSE,message = FALSE}
Annotation_list<- annotateClusters(Genelist_test_cluster, cluster_summary)
```
Here, one could specify
different methods for
pooling p-values
resulting from multiple
terms. In addition, this
pooled p-value is capped
at a `min_pval` (default =
1e-10), making it better
suitable for plotting.
However, users can
select a lower `min_pval`
threshold.
```{r, annotation_adapted, warning = FALSE,message = FALSE}
Annotation_list<- annotateClusters(Genelist_test_cluster, cluster_summary, method = "stouffer", min_pval = 1e-50)
```

## Vizualisation 
The results can be
plotted as a Bubbleplot.
If the enrichment was
done for up-and
downregulation
separately, the plot will
be splitted for up-and
downregulated genes.

```{r, Bubbleplot,  fig.width=8.27, fig.height=7}
plotBubbleplot(Annotation_list)
```
The plot colors can be adjusted, using the `plot_colors` arguement: 

```{r, Bubbleplot_adapted,  fig.width=8.27, fig.height=7}
plotBubbleplot(Annotation_list, plot_colors = list(up = c("#ececec", "#ffb9b1", "#F87060"), down = c("#ececec",
    "#3172CC", "#102542"),  default = c("#ececec", "#3172CC", "#102542")))

```

Additionally, the
results can be plotted
as a Heatmap. If the
enrichment was done for
up-and downregulation
seperatly, the Heatmap
will be spitted for up-and
downregulated genes. The
Heatmap can be further
customized by setting
cluster_rows/columns to
TRUE. The colors can be also changed with the  `plot_colors` argument. 

```{r, Heatmap,  fig.width=8.27, fig.height=7}

plotHeatmap(Annotation_list, rot = 90, annotation_bar = F, column_names_centered = F, plot_colors = list(up = c("#ececec", "#ffb9b1", "#F87060"), down = c("#ececec",
    "#3172CC", "#102542"),  default = c("#ececec", "#3172CC", "#102542")))

```

We can add an `annotation_bar`, that shows the number of terms, contained in each cluster. 


```{r, Heatmap_adapted,  fig.width=8.27, fig.height=7}

plotHeatmap(Annotation_list, rot = 90, annotation_bar = T, column_names_centered = F)

```
